Polarized Cubical Types

### 2014-05-22 ring.tex
thoughts
paths to the rescue

\newcommand{\bdim}[1]{\textsf{dim}(#1)}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
\newcommand{\hast}{$:\mkern -2.5mu:\;$}

@@

Our cubical types $\tau$ include the empty type 0, the unit type 1,
conventional sum and product types, as well as \emph{negative} types:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
The \emph{dimension} of a type is defined as follows:
\[\begin{array}{rcl}
\bdim{0} &=& 0 \\
\bdim{1} &=& 0 \\
\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
\bdim{- \tau} &=& \max(1,\bdim{\tau})
\end{array}\]
The base types have dimension 0. Using negative types but no product types
raises the dimension to 1. This is the situation with the \textbf{Int} or
$\mathcal{G}$ construction. Once product types are used, the dimension can
increase without bounds.

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} \quad= \\
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]

\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (0,0) {$S_1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (0.6,0) {\mm};
\node[right] at (0.6,0) {$S_2$};
\draw[fill] (0.6,0) circle [radius=0.05];
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$};

%%
\node[left] at (2.5,-0.5) {\pp\pp};
\node[below] at (2.5,-0.5) {$S_3$};
\draw[fill] (2.5,-0.5) circle [radius=0.05];
\node[right] at (3.5,-0.5) {\pp\mm};
\node[below] at (3.5,-0.5) {$S_4$};
\draw[fill] (3.5,-0.5) circle [radius=0.05];
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[left] at (2.5,0.5) {\mm\pp};
\node[above] at (2.5,0.5) {$S_5$};
\draw[fill] (2.5,0.5) circle [radius=0.05];
\node[right] at (3.5,0.5) {\mm\mm};
\node[above] at (3.5,0.5) {$S_6$};
\draw[fill] (3.5,0.5) circle [radius=0.05];
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)$\,\mm\pp\pp};
\draw[fill] (7.5,0.75) circle [radius=0.05];
\node[right] at (9.5,0.75) {$(S_2 \times S_4)$\,\mm\pp\mm};
\draw[fill] (9.5,0.75) circle [radius=0.05];
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)$\,\mm\mm\mm};
\draw[fill] (10.2,1.2) circle [radius=0.05];
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)$\,\mm\mm\pp};
\draw[fill] (8.2,1.2) circle [radius=0.05];
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)$\,\pp\pp\pp};
\draw[fill] (7.5,-0.75) circle [radius=0.05];
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)$\,\pp\pp\mm};
\draw[fill] (9.5,-0.75) circle [radius=0.05];
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)$\,\pp\mm\mm};
\draw[fill] (10.2,-0.3) circle [radius=0.05];
\node[left] at (8.2,-0.3) {$(S_1 \times S_5)$\,\pp\mm\pp};
\draw[fill] (8.2,-0.3) circle [radius=0.05];
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{mult}Example of multiplication of two cubical sets.}
\end{figure*}

This point is made precise in the denotation of types which maps a type of
dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
as a binary tree of maximum depth $n$ with nodes of the form
$\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
subspace and $\cubt_2$ (shaded in gray) is the negative subspace along the
first dimension. Each of these subspaces is itself a cube of a lower
dimension. The $0$-dimensional cubes are the conventional denotation of
first-order types as sets. We use $S$ to denote the denotations of such
types. For example, 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which
intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
denotations are $S_1$ and $S_2$ respectively. A full 2-dimensional cube looks
like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
outermost box encodes the sign.

Formally, the denotation of types is as follows:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \star \} \\
\den{\tau_1 + \tau_2} &=& \den{\tau_1} \oplus \den{\tau_2} \\
\den{\tau_1 * \tau_2} &=& \den{\tau_1} \otimes \den{\tau_2} \\
\den{- \tau} &=& \ominus \den{\tau} \\
\\
\noalign{\mbox{where:}\hfill}
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
S \oplus (\nodet{\cubt_1}{\cubt_2}) &=& \nodet{S \oplus \cubt_1}{\cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \oplus S &=& \nodet{\cubt_1 \oplus S}{\cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt} \\
\\
\ominus~S &=& \nodet{\phantom{S}}{S} \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of $0$-dimensional types is the disjoint union as expected. For cubes of
higher dimensions, the subspaces are recursively added. Note that the sum of
1-dimensional types reduces to the sum used in the \textbf{Int} construction.
The definition of negation is natural: it recursively swaps the positive and
negative subspaces. The product of 0-dimensional types is the cartesian
product of sets. For cubes of higher-dimensions $n$ and $m$, the result is of
dimension $(n+m)$. The example in Fig.~\ref{mult} illustrates the idea using
the product of 1-dimensional cube (i.e., a line) with a 2-dimensional cube
(i.e., a square). The result is a 3-dimensional cube as illustrated.

\paragraph*{Note.} 
Our (simple) semantics of types identifies several structurally different
types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
the types are isomorphic. However, in the operational semantics discussed in
the next section, we make the computational content of such type isomorphisms
explicit.

@@

Consider the following simple function on 0-dimensional sum types:
\[\begin{array}{rcl}
\textsf{swapP} &\hast& (\tau_1 + \tau_2) \rightarrow (\tau_2 + \tau_1) \\
\textsf{swapP} (\inl{v}) &=& inr{v} \\
\textsf{swapP} (\inr{v}) &=& inl{v}
\end{code}
Given our setup, this just works $n$-dimensional types. And so do most of the
$\Pi$ combinators for that matter. Only $\eta$ and $\epsilon$ seem to need
thinking. How and why would $1-1$ which is a 1-dimensional line be the same
as the empty type which is a 0-dimensional thing. And how do we generalize
for arbitrary group identities at higher dimensions. We need a mechanism for
cubes with subspaces that ``cancel'' to map to equivalent smaller subcubes.

>>>>

\newcommand{\bdim}[1]{\textsf{dim}(#1)}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
\newcommand{\hast}{:\mkern -2.5mu:\;}

@@

Our cubical types $\tau$ include the empty type 0, the unit type 1,
conventional sum and product types, as well as \emph{negative} types:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
The \emph{dimension} of a type is defined as follows:
\[\begin{array}{rcl}
\bdim{\cdot} &\hast& \tau \rightarrow \mathbb{N} \\
\bdim{0} &=& 0 \\
\bdim{1} &=& 0 \\
\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
\bdim{- \tau} &=& \max(1,\bdim{\tau})
\end{array}\]
The base types have dimension 0. Using negative types but no product types
raises the dimension to 1. This is the situation with the \textbf{Int} or
$\mathcal{G}$ construction. Once product types are used, the dimension can
increase without bounds.

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} \quad= \\
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]
\\
\\
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (0,0) {$S_1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (0.6,0) {\mm};
\node[right] at (0.6,0) {$S_2$};
\draw[fill] (0.6,0) circle [radius=0.05];
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$};

%%
\node[left] at (2.5,-0.5) {\pp\pp};
\node[below] at (2.5,-0.5) {$S_3$};
\draw[fill] (2.5,-0.5) circle [radius=0.05];
\node[right] at (3.5,-0.5) {\pp\mm};
\node[below] at (3.5,-0.5) {$S_4$};
\draw[fill] (3.5,-0.5) circle [radius=0.05];
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[left] at (2.5,0.5) {\mm\pp};
\node[above] at (2.5,0.5) {$S_5$};
\draw[fill] (2.5,0.5) circle [radius=0.05];
\node[right] at (3.5,0.5) {\mm\mm};
\node[above] at (3.5,0.5) {$S_6$};
\draw[fill] (3.5,0.5) circle [radius=0.05];
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)$\,\mm\pp\pp};
\draw[fill] (7.5,0.75) circle [radius=0.05];
\node[right] at (9.5,0.75) {$(S_2 \times S_4)$\,\mm\pp\mm};
\draw[fill] (9.5,0.75) circle [radius=0.05];
\node[above right] at (10.2,1.2) {$(S_2 \times S_6)$\,\mm\mm\mm};
\draw[fill] (10.2,1.2) circle [radius=0.05];
\node[above left] at (8.2,1.2) {$(S_2 \times S_5)$\,\mm\mm\pp};
\draw[fill] (8.2,1.2) circle [radius=0.05];
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)$\,\pp\pp\pp};
\draw[fill] (7.5,-0.75) circle [radius=0.05];
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)$\,\pp\pp\mm};
\draw[fill] (9.5,-0.75) circle [radius=0.05];
\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)$\,\pp\mm\mm};
\draw[fill] (10.2,-0.3) circle [radius=0.05];
\node[left] at (8.2,-0.3) {$(S_1 \times S_5)$\,\pp\mm\pp};
\draw[fill] (8.2,-0.3) circle [radius=0.05];
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

This point is made precise in the denotation of types which maps a type of
dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
as a binary tree of maximum depth~$n$ with nodes of the form
$\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
subspace and $\cubt_2$ (shaded in gray) is the negative subspace along the
first dimension. Each of these subspaces is itself a cube of a lower
dimension. The $0$-dimensional cubes are plain sets representing the
denotation of conventional first-order types. We use $S$ to denote the
denotations of these plain types. A 1-dimensional cube, $\nodet{S_1}{S_2}$,
intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
denotations are $S_1$ and $S_2$ respectively. The type can be visualized as a
``line'' with polarized endpoints connecting the two points $S_1$ and
$S_2$. A full 2-dimensional cube,
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$, intuitively corresponds to
the iterated difference of the appropriate types $(t_1-t_2)-(t_3-t_4)$ where
the product of the successive signs from the outermost box encodes the
sign. The type can be visualized as a ``square'' with polarized corners
connecting the two lines corresponding to $(t_1-t_2)$ and $(t_3-t_4)$. (See
Fig.~\ref{mult} which is further explained after we discuss multiplication
below.)

Formally, the denotation of types is as follows:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \star \} \\
\den{\tau_1 + \tau_2} &=& \den{\tau_1} \oplus \den{\tau_2} \\
\den{\tau_1 * \tau_2} &=& \den{\tau_1} \otimes \den{\tau_2} \\
\den{- \tau} &=& \ominus \den{\tau} \\
\\
\noalign{\mbox{where:}\hfill}
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
S \oplus (\nodet{\cubt_1}{\cubt_2}) &=& \nodet{S \oplus \cubt_1}{\cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \oplus S &=& \nodet{\cubt_1 \oplus S}{\cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt} \\
\\
\ominus~S &=& \nodet{\phantom{S}}{S} \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of $0$-dimensional types is the disjoint union as usual. For cubes of higher
dimensions, the subspaces are recursively added. Note that the sum of
1-dimensional types reduces to the sum used in the \textbf{Int} construction.
The definition of negation is natural: it recursively swaps the positive and
negative subspaces. The product of 0-dimensional types is the cartesian
product of sets. For cubes of higher-dimensions $n$ and $m$, the result is of
dimension $(n+m)$. The example in Fig.~\ref{mult} illustrates the idea using
the product of 1-dimensional cube (i.e., a line) with a 2-dimensional cube
(i.e., a square). The result is a 3-dimensional cube as illustrated.

Our (simple) semantics of types identifies several structurally different
types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
the types are isomorphic. However, in the operational semantics discussed in
the next section, we make the computational content of such type isomorphisms
explicit. Some other isomorphic types like $(t_1*t_2)$ and $(t_2*t_1)$ map to
different cubes and are \emph{not} identified: explicit isomorphisms are
needed to mediate between them. In most cases, the isomorphisms are easy to
define except for the isomorphisms involving the empty type. In particular,
if negative types are to be interpreted as their name suggests, we must have
an isomorphism between $(t-t)$ and the empty type 0. Semantically the former
denotes $\nodet{\cubt}{\cubt}$ and the latter denotes the empty set. Their
denotations are different and there is no way so far to express the fact that
these two cubes should be identified. Intuitively, what is needed is the
ability to contract a line whose endpoints are identical to a single
point. It should then follow that a square where one edge contracts to a
point would itself contract to a line, and a 3-dimensional cube in which one
face contracts to a point would itself contract to a square, etc. This is, of
course, where the ideas of homotopy (type) theory enter the development.

Consider the situation with the simplest instance of the desired isomorphism
which is between $(1-1)$ and the empty type:
\begin{center}
\begin{tikzpicture}
\node[above] at (0,0) {\pp};
\node[left] at (-0.2,0) {$1$};
\draw[fill] (0,0) circle [radius=0.05];
\node[above] at (2.6,0) {\mm};
\node[right] at (2.8,0) {$1$};
\draw[fill] (2.6,0) circle [radius=0.05];
\draw[->,red] (0.1,0) -- (2.5,0);
\draw[->,red] (-0.1,-0.1) arc (135:405:0.2);
\node[below,red] at (0,-0.4) {\textsf{refl}};
\draw[->,red] (2.5,-0.1) arc (135:405:0.2);
\node[below,red] at (2.6,-0.4) {\textsf{refl}};
\node[above] at (1.3,0) {$I$};
\node[right] at (1.3,-0.5) {$q$};
%%
\node[below] at (1.3,-1.5) {$\emptyset$};
\draw[fill] (1.3,-1.5) circle [radius=0.05];
\draw[->,red] (1.4,-1.4) arc (-45:225:0.2);
\node[right,red] at (1.5,-1.3) {\textsf{refl}};
%% 
\draw[->,double,blue] (1.3,-0.1) -- (1.3,-1);
\end{tikzpicture}
\end{center}
The top of the figure is the 1-dimensional cube representing the type $(1-1)$
except that we view the edge connecting the two endpoints as a \emph{path}
$I$ that is initially non-contractible, i.e., it is different from the
identity paths labeled \textsf{refl} that are the trivial paths asserting
that $1$ is contractible to itself. The bottom of the figure is the
0-dimensional cube representing the empty type. There is a trivial path
\textsf{refl} connecting that point to itself. To express the equivalence of
$(1-1)$ to 0, we add a 2-path, i.e. a path between paths, that asserts the
equivalence of the path $I$ with \textsf{refl} on the empty type. That
effectively contracts the two points and collapses them to a single
point. Surprisingly, that is everthing that we need. The extension to higher
dimensions just ``works'' because paths in homotopy type theory have a rich
structure. We explain the details below.

@@

Consider the following simple function on 0-dimensional sum types:
\[\begin{array}{rcl}
\textsf{swapP} &\hast& (\tau_1 + \tau_2) \rightarrow (\tau_2 + \tau_1) \\
\textsf{swapP} (\inl{v}) &=& inr{v} \\
\textsf{swapP} (\inr{v}) &=& inl{v}
\end{array}\]
Given our setup, this just works $n$-dimensional types. And so do most of the
$\Pi$ combinators for that matter. Only $\eta$ and $\epsilon$ seem to need
thinking. How and why would $1-1$ which is a 1-dimensional line be the same
as the empty type which is a 0-dimensional thing. And how do we generalize
for arbitrary group identities at higher dimensions. We need a mechanism for
cubes with subspaces that ``cancel'' to map to equivalent smaller subcubes.
